Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2022-01-03 src/ms.tex
paper edits

\subsection{Does the Sun reside on the long-period pileup?} \label{subsec:thesun}

It is unclear whether or not the Sun is a resident of the long-period pile-up. If one assumes there are no systematic offsets between the \teff scales of the Sun and the spectroscopic catalogs considered here, the Sun's equatorial rotation period places it $\sim$~5--7 days above the long-period pile-up. This raises the question of how the Sun's angular momentum has evolved to its current state, and whether WMB is a generic evolutionary phase.

However, a systematic offset in \teff could place the Sun on the long-period pile-up. To place the Sun on the long-period pile-up would require shifting the spectroscopic temperatures higher by $\sim$50--100~K for the CKS sample, $\sim$100--200~K for the SPOCS sample, or $\sim$200--300~K for the LAMOST sample. However, comparing the Sun to the \citet{Hall2021} asteroseismic sample (Figure~\ref{fig:kde}) seems to support the notion that the Sun may indeed reside on the long-period pile-up, or very close to it, without the need for a \teff shift.

@@

If the Sun is not in fact on the pile-up, one possibility is that the Sun evolved through a phase of weakened magnetic braking in the past and its subsequent period-lengthening was driven by its expansion away from the main sequence \citep[e.g.][]{vanSaders2019}. If this is the case, there is some moderate tension with the ages of pile-up stars in the CKS sample (Figure~\ref{fig:ages}). In the WMB model, hotter stars are expected to depart the pile-up on shorter timescales, yet some CKS stars in the pile-up appear to be both hotter and older than the Sun. It is of course possible that the isochrone ages are inaccurate, particularly for the cooler stars in the sample. If the Sun has indeed evolved through a period of WMB, this implies that the Sun may have a fundamentally different dynamo mechanism to the stars that have not yet transitioned through WMB. This suggests that the Sun's dynamo and consequent manifestations of magnetic activity may not be directly comparable or analogous to the majority of stars with measured Kepler rotation periods (as the majority have not yet transitioned through WMB). This may have implications for the validity of using the Sun as a benchmark calibrator and exemplar in stellar rotation and magnetism models.

It it also possible that the long-period pile-up studied here only applies to stars with masses $\gtrsim$~1~\msun, in which case the Sun's period and age require no explanation. As seen in Figure~\ref{fig:asteroseismic}, it is not clear that a pile-up exists for \teff~$\lesssim$~5800~K, particularly when examining the CKS and \citet{Hall2021} samples (see also \S\ref{subsec:extent}). However, kernel density estimation of the joint \teff--\prot distribution in the \lamostmcq sample, as well as a cross-match between LAMOST and \citet{Santos2021}, suggests the pile-up may exist at \teff as low as 5500~K or even 5000~K. {\color{red} Furthermore...} Further work with a larger sample size may reveal if the pile-up does indeed extend to temperatures as cool as the Sun's.

Another question posed by the Sun in relation to the long-period pile-up is how such a feature can appear so well-defined given the spread of periods observed on the Sun due to differential rotation (Figure~\ref{fig:surveys}). If other Sun-like stars have a similar degree of differential rotation, one might expect the long-period edge to be blurred out. One possible explanation is that starspots at higher latitudes produce smaller variations in light curves as they are in view or out of view for most of a rotation cycle for some random inclination \citep[e.g.][]{Luger2021a}. Forward modeling simulations of the Kepler field including the effects of surface differential rotation may be able to answer this question.

@@

    \item The long-period pile-up is reasonably well-described by a constant Rossby number, with a critical value of \rocritfinal, in the \teff range of $\approx$~5000-6250~K. The pile-up is also  populated by stars with a wide range of isochrone ages (\agerange). A pile-up of stars with a constant Rossby number and a broad range of ages is a prediction of the WMB model of \citet{vanSaders2016, vanSaders2019}. The precise value of \rocrit is highly sensitive to \teff scale shifts between observational data and the models used to compute \taucz.

>>>>

\subsection{Does the Sun reside on the long-period pileup?} \label{subsec:thesun}
{\color{red}

It is unclear whether or not the Sun is a resident of the long-period pile-up. If one assumes there are no systematic offsets between the \teff scales of the Sun and the spectroscopic catalogs considered here, the Sun's equatorial rotation period places it $\sim$5--7 days above the long-period pile-up. This raises the question of how the Sun's angular momentum has evolved to its current state, and whether WMB is a generic evolutionary phase.

One possibility is that the Sun evolved through a phase of weakened magnetic braking in the past and its subsequent period-lengthening was driven by its expansion away from the main sequence \citep[e.g.][]{vanSaders2019}. However, this scenario is considered implausible as the change in moment of inertia for the Sun over its lifetime is not enough to produce the 5--7 day gap between the Sun and the observed long-period pile-up.

It it also possible that the long-period pile-up studied here only applies to stars with masses $\gtrsim$~1~\msun, in which case the Sun's period and age require no explanation. However, in this scenario we would have no explanation for why hotter stars pile-up while cooler stars do not, especially given that the pile-up appears to occur at low enough Ro that the cooler stars should appear to pile up as well. As seen in Figure~\ref{fig:asteroseismic}, it is not clear that a pile-up exists for \teff~$\lesssim$~5800~K, particularly when examining the CKS and \citet{Hall2021} samples (see also \S\ref{subsec:extent}). However, kernel density estimation of the joint \teff--\prot distribution in the \lamostmcq sample, as well as a cross-match between LAMOST and \citet{Santos2021}, suggests the pile-up may exist at \teff as low as 5500~K or even 5000~K. We thus consider the explanation that WMB only operates for $\gtrsim$~1~\msun stars unlikely, and suggest that further work with a larger sample size could reveal the true \teff range of the pile-up more clearly.

We consider the most likely explanation for the observed gap between the Sun and the long-period pile-up to be systematic offsets in \teff. To place the Sun on the long-period pile-up would require shifting the spectroscopic temperatures higher by $\sim$50--100~K for the CKS sample, $\sim$100--200~K for the SPOCS sample, or $\sim$200--300~K for the LAMOST sample. Notably, comparing the Sun to the \citet{Hall2021} asteroseismic sample (Figure~\ref{fig:kde}) seems to support the notion that the Sun may indeed reside on the long-period pile-up, or very close to it, without the need for a \teff shift. Furthermore, as discussed in \S\ref{subsec:longperiod}, the \citet{Hall2021} sample is less subject to detection bias which acts to censor stars with higher Ro and lower variability amplitudes in the samples considered here. As shown by \citet{Aigrain2015}, the pipelines used to measure rotation periods from Kepler data would not be guaranteed to detect the solar rotation period. We conclude that the Sun likely does reside on the long-period pile-up, but that this is obfuscated in the present samples by temperature offsets as well as potentially detection biases and small sample sizes.
}

%If this is the case, there is some moderate tension with the ages of pile-up stars in the CKS sample (Figure~\ref{fig:ages}). In the WMB model, hotter stars are expected to depart the pile-up on shorter timescales, yet some CKS stars in the pile-up appear to be both hotter and older than the Sun. It is of course possible that the isochrone ages are inaccurate, particularly for the cooler stars in the sample. If the Sun has indeed evolved through a period of WMB, this implies that the Sun may have a fundamentally different dynamo mechanism to the stars that have not yet transitioned through WMB. This suggests that the Sun's dynamo and consequent manifestations of magnetic activity may not be directly comparable or analogous to the majority of stars with measured Kepler rotation periods (as the majority have not yet transitioned through WMB). This may have implications for the validity of using the Sun as a benchmark calibrator and exemplar in stellar rotation and magnetism models.

@@

%Another question posed by the Sun in relation to the long-period pile-up is how such a feature can appear so well-defined given the spread of periods observed on the Sun due to differential rotation (Figure~\ref{fig:surveys}). If other Sun-like stars have a similar degree of differential rotation, one might expect the long-period edge to be blurred out. One possible explanation is that starspots at higher latitudes produce smaller variations in light curves as they are in view or out of view for most of a rotation cycle for some random inclination \citep[e.g.][]{Luger2021a}. Forward modeling simulations of the Kepler field including the effects of surface differential rotation may be able to answer this question.

@@

    \item The long-period pile-up is well-described by a constant Rossby number, with a critical value of \rocritfinal, in the \teff range of $\approx$~5000-6250~K. The pile-up is also populated by stars with a wide range of isochrone ages (\agerange). A pile-up of stars with a constant Rossby number and a broad range of ages is a prediction of the WMB model of \citet{vanSaders2016, vanSaders2019}. The precise value of \rocrit is sensitive to \teff scale shifts between observational data and the models used to compute \taucz.
